### 2017-06-08 2017/NACO/paper.tex
Исправлены замечания Гергеля

% It is required to enter 2010 MSC.
\subjclass{Primary: 90C26; Secondary: 65Y05.}
% Please provide minimum  5 keywords.
 \keywords{Global optimization, information-statistical theory, parallel computing, high-performance
computer systems,  supercomputer technologies.}

@@

\centerline{\scshape Victor Gergel, Konstantin  Barkalov$^*$ and Alexander Sysoyev}
\medskip
{\footnotesize
 % please put the address of the second  and third author
 \centerline{Lobachevsky State University of Nizhny Novgorod}
   \centerline{23, Gagarin Avenue}
   \centerline{Nizhny Novgorod, Russia}
}

@@

the criterion to be optimized has several local optima within the search domain,
which have different values. The existence of several local optima essentially makes

@@

a reasonable time by employing parallel computations on modern supercomputer systems only.

@@

The general state of the art in the field of global optimization has been presented in a
number of key monographs  \cite{floudasPardGO}, \cite{horstTuyGO}, \cite{locatelliSchoenGO}, \cite{pinterGO}, \cite{strSergGO}, \cite{zilinskTornGO}, \cite{zhigljavskyRandGO}, etc.

@@

problems as well as of the systems of the nonlinear equations. The system includes

@@

\item IOSO (Indirect Optimization on the basis of Self-Organization) \cite{iosoDescription} is oriented on solving a wide class of the optimization problems including the global optimization ones. The system is widely used to solve the applied problems in various application fields. There is a version of the system for exciting on the parallel computational systems. The system is a commercial product, but it is available for the trial usage.

@@

In this paper, the core class\footnote{In general, the Globalizer can be applied for solving
multicriterial multiextremal multidimensional optimization problems with nonlinear constraints.}
of optimization problems, which can be solved using

@@

  \varphi(y^*)=\min\{\varphi(y):y\in D\} \\

@@

It is supposed, that the objective function \(\varphi(y)\) satisfies the Lipschitz condition

@@

of the next trial points is determined by using the search information ((the preceding

@@

for global optimization methods since it implies a complex analysis of a large amount

@@

\varphi(y(x^*))=\min\{\varphi(y(x)):x\in [0,1]\}

@@

|\varphi(y(x_1))-\varphi(y(x_2))|\leq H{|x_1-x_2|}^{\frac{1}{N}},x_1,x_2\in[0,1]

@@

where the constant H is defined by the relation \(H=4L\sqrt{N}\), \(L\) is the Lipschitz

@@

presented in \cite{strSergGO}. As an illustration, an approximation of the Peano curve

@@

    \caption{A Peano curve approximation for the third level}

@@

\par
The computational scheme obtained as a result of the dimensionality reduction consists of the following
(see Figure \ref{fig:peanoCUsage}):
\begin{itemize}
  \item The optimization algorithm performs the minimization of the reduced one-dimensional
  function \(\varphi(y(x))\) from (\ref{eq:oneDimTask}),
  \item After determining the next trial point \(x\), a multidimensional image \(y\) in the mapping \(y(x)\) is calculated,
  \item The value of the initial multidimensional function \(\varphi(y)\) is calculated at the multidimensional point \(y\in D\),
  \item The calculated value \(z=\varphi(y)\) is used further as the value of the reduced one-dimensional function \(\varphi(y(x))\) at the point \(x\).
\end{itemize}

@@

\section{Parallel Computations for Solving Global Optimization Problems}
\label{sec:parallel}
Let us consider the parallelization approaches used widely in the theoretical and
practical applications of parallel computing within the context of global optimization problems:
  \item The distribution of the search domain \(D\) among the available computational units (data parallelization scheme). This approach is insufficient in the case of optimization problems since in such computations the subdomain, which contain the sought global minimum, will only be processed by one processor, and, therefore, all the remaining processors would perform the excess computations.
  \item The parallelization of the optimization algorithms (task parallelization scheme). This approach is also insufficient since the direct computational costs for executing the optimization algorithms are relatively low (the majority of computations are represented by the calculations of the optimized function values due to the initial assumption of considerable computational costs of such calculations).
  \item The parallelization of the computations executed in order to obtain the values of the optimized function. This approach is applicable since the most time-consuming part of the global optimization will be parallelized. However, this method is not featured by the generality (the development of parallelization methods is to be performed every time from the very beginning while solving each particular optimization problem).
Within the framework of the information-statistical theory, a general parallelization
approach for solving the global optimization problems has been proposed [\cite{stronginGergelBarkalovParGO}, \cite{strSergGO} --- \textit{the
parallel computations are provided by means of simultaneous computing the values of the
minimized function \(\varphi(y)\) at several different points within the search domain \(D\)}.
This approach provides the parallelization for the most time-consuming part of the optimization computations.
The global optimization algorithms implemented in Globalizer will be described step-by-step.
In Subsection \ref{subsec:corepar}, the core serial Multidimensional Algorithm of Global
Search (MAGS) is presented. In Subsection \ref{subsec:sharedpar}, a parallel generalization
of the MAGS algorithm for the parallel computations on the multiprocessor computational
nodes with shared memory is described. In Subsection \ref{subsec:distribpar}, the scheme
for the parallel computations on the high-performance computational systems with distributed memory is given.
\subsection{Core Multidimensional Generalized Algorithm of Global Search}
\label{subsec:corepar}
The information-statistical theory of global optimization formulated in \cite{strongin1978}, \cite{strSergGO} has
served as a basis for the development of a large number of efficient multiextremal optimization
methods \cite{barkalovGergel2014}, \cite{Pizzuti}, \cite{gergel1996}, \cite{gergel1997}, \cite{grishaginStrongin1984}, \cite{sergeyev1995}, \cite{sergeyev1999}, \cite{sergeyevGrishagin2001}, \cite{sergeyevStronginLera2013}, \cite{Famularo2001}.
The optimization methods applied in Globalizer are based on the MAGS method, which can be presented as follows --- see \cite{strongin1978}, \cite{strSergGO}.
Let us introduce a simpler notation for the optimization problem being solved
\label{eq:oneDimFunc}
f(x) = \varphi(y(x)):x\in [0,1].
The initial iteration of the algorithm is performed at an arbitrary point \(x^1(0,1)\).
Then, let us suppose that \(k\), \(k\ge 1\), optimization iterations have been completed already.
The selection of the trial point \(k+1\) for the next iteration is performed according to the following rules.
\textit{Rule 1}. Renumber the points of the preceding trials by the lower indices in order of increasing value of coordinates
  \label{step1}
0=x_0<x_1<...<x_{k+1}=1.
the points \(x_0\), \(x_k\) were introduced additionally for the convenience of further
explanation, the values of the minimized function \(z_0\), \(z_k\) at these points are undefined.
\textit{Rule 2}. Compute current estimate of the H{\"o}lder constant \(H\) from (\ref{eq:holder})
\begin{equation} \label{step2}
m=\max_{1\leq i\leq k}\frac{|z_i-z_{i-1}|}{\rho_i}, \;
M = \left\{
   \begin{array}{lr}
     rm, & m > 0,\\
     1, & m = 0,
  \right.
as the maximum of the relative differences of the minimized function values on the
set of previously executed trial points \(x_i,1\leq i\leq k\) from (\ref{step1}).
Hereafter, \(\rho_i=(x_i-x_{i-1})^\frac{1}{N},1\leq i\leq k+1\). The
constant \(r\), \(r>1\), is the reliability parameter of the algorithm.
\textit{Rule 3}. Compute the characteristics \(R(i)\) for each interval \((x_{i-1},x_i),1\leq i\leq k+1\), where
\[
R(i)=2\rho_i-4\frac{z_i}{M},\quad i=1,
\]
\begin{equation} \label{step3}
R(i)=\rho_i+\frac{(z_i-z_{i-1})^2}{M^2\rho_i}-2\frac{z_i+z_{i-1}}{M},\quad 1<i<k+1, \\
\[
R(i)=2\rho_{i}-4\frac{z_{i-1}}{M},\quad i=k+1.
\]

\par
\textit{Rule 4}. Determine the interval with the maximum characteristic
\begin{equation} \label{step4}
R(t)=\max_{1\leq i \leq k+1}R(i).
\end{equation}
\par
\textit{Rule 5}. Execute a new trial at the point \(x^{k+1}\) located within the interval
with the maximum characteristic from (\ref{step4})
\begin{equation} \label{step5}
  x_{k+1}=\frac{x_t+x_{t-1}}{2}-\mathrm{sign}(z_{t}-z_{t-1})\frac{1}{2r}\left[\frac{|z_{t}-z_{t-1}|}{M}\right]^N,\; \textrm{ if } 1<t<k+1,
\end{equation}
\[
  x_{k+1}=\frac{x_t+x_{t-1}}{2},\; \textrm{ if } t=1 \textrm{ or } t=k+1.
\]

\par
The stop condition, which terminated the trials, is defined by the inequality
\begin{equation}
  \label{eq:stop_1}
\rho_t<\varepsilon
\end{equation}
for the interval with the maximum characteristic from (\ref{step4}) and \(\varepsilon >0\) is the predefined
accuracy of the optimization problem solution. If the stop condition is not satisfied,
the index \(k\) is incremented by 1, and new global optimization iteration is executed.
\par
In order to explain the algorithm presented above, let us note the following.
The characteristics \(R(i), 1\leq i\leq k+1\), calculated according to (\ref{step3}) could
be interpreted as some measures of importance of the intervals with respect to the
expected location of the global minimum point. Thus, the rules (\ref{step4}) and (\ref{step5}) for selecting
the interval for executing the next trial become more clear --- the point of every next
global optimization iteration is selected within the interval, where the global minimum
point can be found most likely.
\par
The convergence conditions for the described algorithm are given, for example, in \cite{strSergGO}.

\subsection{Parallel Computations for Systems with Shared Memory}
\label{subsec:sharedpar}
Modern supercomputer computational systems consist of plenty of computational nodes,
which include several multicore processors. In this case, random access memory is
shared among the computational nodes --- the content of any memory location can be read
(written) by any computational core at any arbitrary moment. In most cases, shared memory is
uniform --- the time characteristics for accessing the memory are the same for all
computational cores and for all memory locations.
\par
The following approach can be applied to provide the parallel computations with the MAGS method.
As it was mentioned above, the characteristics \(R(i),1\leq i\leq k+1\) calculated
according to (\ref{step3}) can be interpreted as some measures of importance of the
intervals with respect to the expected location of the global minimum point.
Following this understanding, every next global optimization iteration is executed within
the most important interval with the maximum characteristic. As a result, it becomes
evident how to select the other intervals for simultaneous computations of the minimized
function values at several different points within the search domain --- these ones
should be the intervals with the next magnitudes of characteristics.
\par
The computational scheme of the Parallel Multidimensional Algorithm of Global Search (PMAGS)
for the computational systems with shared memory can be developed based on the approach
described above. This scheme is almost identical to the MAGS one --- the differences
consist just in the following set of rules.
\par
\textit{Rule 4'}. Arrange the characteristics of the intervals computed according to (\ref{step3}) in the decreasing order
\begin{equation}
\label{step4par}
R(t_1)\geq R(t_2)\geq \dots \geq R(t_{k})\geq R(t_{k+1})
\end{equation}
and select \(p\) intervals with the indices \(t_j,1\leq j\leq p\), with the maximum
values of characteristics (\(p\) is the number of processors/cores used for the parallel computations).
\par
\textit{Rule 5'}. Execute new trials at the points \(x_{k+j},1\leq j\leq p\) located
in the intervals with the maximum characteristics from (\ref{step4par})

\begin{equation} \label{step5par}
x_{k+j}=\frac{x_{t_j}+x_{t_j-1}}{2}-\mathrm{sign}(z_{t_j}-z_{t_j-1})\frac{1}{2r}\left[\frac{|z_{t_j}-z_{t_j-1}|}{M}\right]^N,\; \textrm{ if } 1<t_j<k+1,
\end{equation}
\[
x_{k+j}=\frac{x_{t_j}+x_{t_j-1}}{2},\; \textrm{ if } t_j=1 \textrm{ or } t_j=k+1.
\]
\par
The stop condition (\ref{eq:stop_1}) should be checked for all intervals from (\ref{step4par}),
where the scheduled trials are executed
\begin{equation}
  \label{eq:stop}
\rho_{t_j}<\varepsilon,1\leq j\leq p.
\end{equation}
If the stop condition is satisfied at least for one interval \(t_j, 1\le j\le p\), the calculations
are terminated. Otherwise, the index \(k\) is incremented by \(p\), and new global optimization iteration is executed.
\par
The convergence conditions for the developed parallel algorithm are considered
in \cite{stronginGergelBarkalovParGO}, \cite{strSergGO}. Thus, in particular, when the convergence conditions are
satisfied, the parallel computations are non-redundant as compared to the serial method
when using up to \(2^N\) processors/cores (\(N\) is the dimensionality of the global optimization problem).

@@

problems of the kind (\ref{eq:oneDimTask}):

@@

        x_i':y'=y^i(x_i'),1\le i\le s,

@@

The possibility of asynchronous data transfer (the computational nodes process the received
data upon receipt only) is an important feature of the proposed parallel computation scheme.

@@

\par
The architecture of the Globalizer system is presented in Figure \ref{fig:globalizerScheme}.
The system components are:
\begin{itemize}
  \item Block 0 consists of the procedures for computing the function values
  (criteria and constraints) of the optimization problem being solved; it is an external
  block with a predefined interface with respect to the Globalizer system and is provided by the user.
  \item Blocks 1-4 form the optimization subsystem and solve the global optimization
  problems (Block 1), nonlinear programming ones (Block 2), multicriterial optimization problems
  (Block 3), and general decision making problems (Block 4). The interaction between these
  components is organized in a hierarchical order --- the decision making problems are
  solved using the multicriterial optimization block, which, in turn, uses the nonlinear programming block, and so on.
  \item Block 5 is a subsystem for accumulating and processing the search information; it is one of the main subsystems – the size of search information may appear to be quite large and the efficiency of the global optimization methods depends on how completely the available search data are utilized.
  \item Block 6 contains the dimensionality reduction procedures based on the Peano space-filling curves; the reduced (one-dimensional) optimization problems are solved by the optimization blocks. This block also provides the interaction between the optimization blocks and the multidimensional optimization problem being solved.
  \item Block 7 provides tuning the parallel computation schemes in the Globalizer system subject to the employed computational system architecture (the number of the computational cores, the availability of shared and/or distributed memory, the availability of computational accelerators, etc.) and the applied global optimization methods.
  \item Block 8 is responsible for managing the parallel processes when performing the global optimization (determining the optimal configuration of parallel processes, distributing the processes between the computational unit, balancing the computational loads, etc.).
  \item Block 9 is a managing subsystem, which controls the parallel computations when solving the global optimization problems.
  \item Block 10 provides the dialog interaction with the users for stating the optimization problem, adjusting the system parameters (if necessary), and visualizing and presenting the global optimization results.
  \item Block 11 is a set of tools for visualizing and presenting the global optimization results; the availability of the tools for visual demonstration of the computational results enables the user to perform an efficient control over the global optimization process.
\end{itemize}
\par
The Globalizer system comes in two different versions:
\begin{itemize}
  \item Globalizer Pro is a research version for making various large-scale numerical experiments on the high-performance computational systems to investigate the new parallel global optimization methods being developed.
  \item Globalizer Lite is a limited version for solving the global optimization problems with a medium complexity level, and is accessible free of charge.
\end{itemize}

\section{Results of Numerical Experiments}
\label{sec:experiments}
The numerical experiments were carried out using the Lobachevsky supercomputer at
the State University of Nizhni Novgorod (the operation system – CentOS 6.4, the
supercomputer management system – SLURM). Each supercomputer node had 2 Intel Sandy Bridge
E5-2660 processors, 2.2 GHz, with 64 GB RAM. Each processor had 8 cores (i. e. total
16 CPU cores were available at each node). In order to generate the executable program code, the Intel C++ 14.0.2 compiler was used.
\par
A number of computational nodes had Intel Xeon Phi 5110P processors, each containing 60 cores
(total 240 threads). In addition, the supercomputer included the computational nodes equipped
with two NVIDIA Kepler K20X GPUs, each providing 14 stream multiprocessors (total 2688 CUDA-cores). In order to generate the executable code, the CUDA Toolkit 6.0 was used.
\par
In the performed experiments the optimization problems generated by the GKLS-generator
\cite{gavianoKvasovLeraSergeev2003} were used. This generator allows obtaining the multiextremal optimization
problems with \textit{a priori} known properties: the number of local minima, the sizes of the attractors of these ones, the global minimum point, the function value at this point, etc. In order to simulate the computational costs inherent to the applied optimization problems, in all experiments computing the objective function was intentionally complicated by the auxiliary computations not altering the form of the function and the position of its minima.
\par
The numerical experiments were carried out first using the ExaMin system and then were reproduced using the Globalizer system. In all the executed experiments, 100 global optimization problems generated randomly by the GKLS-generator were solved. The results of experiments were averaged over the number of the solved optimization problems.

\subsection{Comparison with over methods}
The results of numerical comparison of three sequential algorithms --- DIRECT
\cite{Jones}, DIRECT\textit{l} \cite{Gablonsky}, and MAGS from Section \ref{subsec:corepar} --- are presented below.

The numerical comparison was carried out by using the function classes Simple and Hard
of the dimensions 4 and 5. The results of numerical experiments for DIRECT and
DIRECT\textit{l} are taken from \cite{sergeyevKvasov2006}. According to \cite{sergeyevKvasov2006}, the global minimizer \(y^*\)
was considered to be found if the algorithm generated a trial point \(y^k\) inside
a hyperinterval with a vertex \(y^*\) and
\begin{displaymath}
    |y^k(i)-y^*(i)|\le \sqrt[N]{\Delta}(b(i)-a(i)), i\le i\le N
\end{displaymath}
where \(N\) is the problem dimensionality, a and b are the borders of the search domain \(D\)
\footnote{In the case of MAGS, the stop condition was different slightly --- the
global minimizer \(y^*\) was considered to be found, if MAGS generates a trial
point \(y^k\)  in the \(\delta\)-vicinity of the global minimum, i. e. \(||y^k-y^*||\le\delta\).
The size of the vicinity was selected  as \(\delta=\left\|b-a\right\|\sqrt[N]{\Delta}\).}.

>>>>

% It is required to enter 2010 MSC.
\subjclass{Primary: 90C26; Secondary: 65Y05.}
% Please provide minimum  5 keywords.
 \keywords{Global optimization, information-statistical theory, parallel computing, high-performance
computational systems, supercomputing technologies.}

@@

\centerline{\scshape Victor Gergel, Konstantin  Barkalov$^*$ and Alexander Sysoyev}
\medskip
{\footnotesize
 % please put the address of the second  and third author
 \centerline{Lobachevsky State University of Nizhny Novgorod}
   \centerline{23, Gagarin Avenue, Nizhni Novgorod, Russia}
}

@@

the optimization criterion can have several local optima within the search domain. 
The existence of several local optima essentially makes

@@

a reasonable time by employing parallel computations on modern supercomputing systems only.

@@

The general state of the art in the field of global optimization is presented in a
number of key monographs  \cite{floudasPardGO}, \cite{horstTuyGO}, \cite{locatelliSchoenGO}, \cite{pinterGO}, \cite{strSergGO}, \cite{zilinskTornGO}, \cite{zhigljavskyRandGO}.

@@

problems as well as the systems of the nonlinear equations. The system includes

@@

\item IOSO (Indirect Optimization on the basis of Self-Organization) \cite{iosoDescription} is oriented on solving a wide class of the optimization problems including the global optimization ones. The system is widely used to solve the applied problems in various application fields. There is a version of the system for executing on the parallel computational systems. The system is a commercial product, but it is available for the trial usage.

@@

In this paper, the core class of optimization problems, which can be solved using

@@

  \varphi(y^*)=\min\{\varphi(y):y\in D\}, \\

@@

with the given boundary vectors  $a$ and  $b$. It is supposed, that the objective function \(\varphi(y)\) satisfies the Lipschitz condition

@@

of the next trial points is determined by using the search information (the preceding

@@

of global optimization methods since it implies a complex analysis of a large amount

@@

\varphi(y(x^*))=\min\{\varphi(y(x)):x\in [0,1]\}.

@@

|\varphi(y(x_1))-\varphi(y(x_2))|\leq H{|x_1-x_2|}^{\frac{1}{N}}, x_1,x_2\in[0,1],

@@

where the constant $H$ is defined by the relation \(H=2L\sqrt{N+3}\), \(L\) is the Lipschitz

@@

given in \cite{strSergGO}. As an illustration, an approximation of the Peano curve

@@

    \caption{A Peano curve approximation for the third density}

@@

\par
The computational scheme obtained as a result of the dimensionality reduction consists of the following
(see Figure \ref{fig:peanoCUsage}):
\begin{itemize}
  \item The optimization algorithm performs the minimization of the reduced one-dimensional
  function \(\varphi(y(x))\) from (\ref{eq:oneDimTask}),
  \item After determining the next trial point \(x\), a multidimensional image \(y\) is calculated by using the
mapping \(y(x)\),
  \item The value of the initial multidimensional function \(\varphi(y)\) is calculated at the point \(y\in D\),
  \item The calculated value \(z=\varphi(y)\) is used further as the value of the reduced one-dimensional function \(\varphi(y(x))\) at the point \(x\).
\end{itemize}

@@

\section{Parallel Computations for Solving Global Optimization Problems}
\label{sec:parallel}
Let us consider the parallelization approaches used widely in the theoretical and
practical applications of parallel computing within the context of global optimization problems:
\begin{itemize}
  \item The distribution of the search domain \(D\) among the available computational units (data parallelization scheme). This approach is insufficient in the case of optimization problems since in such computations the subdomain, which contain the sought global minimum, will only be processed by one processor, and, therefore, all the remaining processors would perform the excess computations.
  \item The parallelization of the optimization algorithms (task parallelization scheme). This approach is also insufficient since the direct computational costs for executing the optimization algorithms are relatively low (the majority of computations are represented by the calculations of the optimized function values due to the initial assumption of considerable computational costs of such calculations).
  \item The parallelization of the computations executed in order to obtain the values of the optimized function. This approach is applicable since the most time-consuming part of the global optimization will be parallelized. However, this method is not featured by the generality (the development of parallelization methods is to be performed every time from the very beginning while solving each particular optimization problem).
\end{itemize}
\par
Within the framework of the information-statistical theory, a general parallelization
approach for solving the global optimization problems has been proposed \cite{stronginGergelBarkalovParGO}, \cite{strSergGO} --- \textit{the parallel computations are provided by means of simultaneous computing the values of the
minimized function \(\varphi(y)\) at several different points within the search domain \(D\)}.
This approach provides the parallelization for the most time-consuming part of the optimization computations.
\par
The global optimization algorithms implemented in Globalizer will be described step-by-step.
In Subsection \ref{subsec:corepar}, the core serial Multidimensional Algorithm of Global
Search (MAGS) is presented. In Subsection \ref{subsec:sharedpar}, a parallel generalization
of the MAGS algorithm for the parallel computations on the multiprocessor computational
nodes with shared memory is described. In Subsection \ref{subsec:distribpar}, the scheme
for the parallel computations on the high-performance computational systems with distributed memory is given.

\subsection{Core Multidimensional Algorithm of Global Search}
\label{subsec:corepar}
The information-statistical theory of global optimization formulated in \cite{strongin1978}, \cite{strSergGO} has
served as a basis for the development of a large number of efficient multiextremal optimization
methods \cite{barkalovGergel2014}, \cite{gergel1996}, \cite{gergel1997}, \cite{grishaginStrongin1984}, \cite{Pizzuti}, \cite{sergeyev1995}, \cite{sergeyev1999}, \cite{sergeyevGrishagin2001}, \cite{sergeyevStronginLera2013}, \cite{Famularo2001}.
\par
The optimization methods applied in Globalizer are based on the MAGS method, which can be presented as follows --- see \cite{strongin1978}, \cite{strSergGO}.
\par
Let us introduce a simpler notation for the optimization problem being solved
\begin{equation}
\label{eq:oneDimFunc}
f(x) = \varphi(y(x)):x\in [0,1].
\end{equation}
\par
The initial iteration of the algorithm is performed at an arbitrary point \mbox{\(x^1\in(0,1)\)}.
Then, let us suppose that \(k\), \(k\ge 1\), optimization iterations have been completed already.
The selection of the trial point \(x^{k+1}\) for the next iteration is performed according to the following rules.
\par
\textit{Rule 1}. Renumber the points of the preceding trials by the lower indices in order of increasing value of coordinates
\begin{equation}
  \label{step1}
0=x_0<x_1<...<x_{k+1}=1,
\end{equation}
the points \(x_0\), \(x_k\) were introduced additionally for the convenience of further
explanation, the values of the minimized function \(z_0\), \(z_k\) at these points are undefined.
\par
\textit{Rule 2}. Compute current estimate of the H{\"o}lder constant \(H\) from (\ref{eq:holder})
\begin{equation} \label{step2}
M=\max_{1\leq i\leq k}\frac{|z_i-z_{i-1}|}{\rho_i}, \;
m = \left\{
   \begin{array}{lr}
     rM, & M > 0,\\
     1, & M = 0,
   \end{array}
  \right.
\end{equation}
as the maximum of the relative differences of the minimized function values on the
set of previously executed trial points \(x_i,1\leq i\leq k\) from (\ref{step1}).
Hereafter, \(\rho_i=(x_i-x_{i-1})^\frac{1}{N},1\leq i\leq k+1\). The
constant \(r\), \(r>1\), is the reliability parameter of the algorithm.
\par
\textit{Rule 3}. Compute the characteristics \(R(i)\) for each interval \((x_{i-1},x_i),1\leq i\leq k+1\), where
\[
R(i)=2\rho_i-4\frac{z_i}{m},\quad i=1,
\]
\begin{equation} \label{step3}
R(i)=\rho_i+\frac{(z_i-z_{i-1})^2}{m^2\rho_i}-2\frac{z_i+z_{i-1}}{m},\quad 1<i<k+1, \\
\end{equation}
\[
R(i)=2\rho_{i}-4\frac{z_{i-1}}{m},\quad i=k+1.
\]

\par
\textit{Rule 4}. Determine the interval with the maximum characteristic
\begin{equation} \label{step4}
R(t)=\max_{1\leq i \leq k+1}R(i).
\end{equation}
\par
\textit{Rule 5}. Execute a new trial at the point \(x^{k+1}\) located within the interval
with the maximum characteristic from (\ref{step4})
\begin{equation} \label{step5}
  x^{k+1}=\frac{x_t+x_{t-1}}{2}-\mathrm{sign}(z_{t}-z_{t-1})\frac{1}{2r}\left[\frac{|z_{t}-z_{t-1}|}{M}\right]^N,\; \textrm{ if } 1<t<k+1,
\end{equation}
\[
  x^{k+1}=\frac{x_t+x_{t-1}}{2},\; \textrm{ if } t=1 \textrm{ or } t=k+1.
\]

\par
The stopping condition, which terminated the trials, is defined by the inequality
\begin{equation}
  \label{eq:stop_1}
\rho_t<\varepsilon
\end{equation}
for the interval with the maximum characteristic from (\ref{step4}) and \(\varepsilon >0\) is the predefined
accuracy of the optimization problem solution. If the stopping condition is not satisfied,
the index \(k\) is incremented by 1, and the new global optimization iteration is executed.
\par
In order to explain the algorithm presented above, let us note the following.
The characteristics \(R(i), 1\leq i\leq k+1\), calculated according to (\ref{step3}) could
be interpreted as some measures of importance of the intervals with respect to the
expected location of the global minimum point. Thus, the rules (\ref{step4}) and (\ref{step5}) for selecting
the interval for executing the next trial become more clear --- the point of every next
global optimization iteration is selected within the interval, where the global minimum
point can be found most likely.
\par
The convergence conditions of the described algorithm are given, for example, in \cite{strSergGO}.

\subsection{Parallel Computations for Systems with Shared Memory}
\label{subsec:sharedpar}
Modern supercomputer computational systems consist of plenty of computational nodes,
which include several multicore processors. In this case, random access memory
in the computational nodes is shared --- the content of any memory location can be read
(written) by any computational core at any arbitrary moment. In most cases, shared memory is
uniform --- the time characteristics for accessing the memory are the same for all
computational cores and for all memory locations.
\par
The following approach can be applied to provide the parallel computations with the MAGS method.
As it was mentioned above, the characteristics \(R(i),1\leq i\leq k+1\) calculated
according to (\ref{step3}) can be interpreted as some measures of importance of the
intervals with respect to the expected location of the global minimum point.
Following this understanding, every next global optimization iteration is executed within
the most important interval with the maximum characteristic. As a result, it becomes
evident how to select the other intervals for simultaneous computations of the minimized
function values at several different points within the search domain --- these ones
should be the intervals with the next magnitudes of characteristics.
\par
The computational scheme of the Parallel Multidimensional Algorithm of Global Search (PMAGS)
for the computational systems with shared memory can be developed based on the approach
described above. This scheme is almost identical to the MAGS one --- the differences
consist just in the following set of rules.
\par
\textit{Rule 4\'}. Arrange the characteristics of the intervals computed according to (\ref{step3}) in the decreasing order
\begin{equation}
\label{step4par}
R(t_1)\geq R(t_2)\geq \dots \geq R(t_{k})\geq R(t_{k+1})
\end{equation}
and select \(p\) intervals with the indices \(t_j,1\leq j\leq p\), with the maximum
values of characteristics (\(p\) is the number of processors/cores used for the parallel computations).
\par
\textit{Rule 5\'}. Execute new trials at the points \(x^{k+j},1\leq j\leq p\) located
in the intervals with the maximum characteristics from (\ref{step4par})

\begin{equation} \label{step5par}
x^{k+j}=\frac{x_{t_j}+x_{t_j-1}}{2}-\mathrm{sign}(z_{t_j}-z_{t_j-1})\frac{1}{2r}\left[\frac{|z_{t_j}-z_{t_j-1}|}{M}\right]^N,\; \textrm{ if } 1<t_j<k+1,
\end{equation}
\[
x^{k+j}=\frac{x_{t_j}+x_{t_j-1}}{2},\; \textrm{ if } t_j=1 \textrm{ or } t_j=k+1.
\]
\par
The stopping condition (\ref{eq:stop_1}) should be checked for all intervals from (\ref{step4par}),
where the scheduled trials are executed
\begin{equation}
  \label{eq:stop}
\rho_{t_j}<\varepsilon,1\leq j\leq p.
\end{equation}
If the stopping condition is satisfied at least for one interval \(t_j, 1\le j\le p\), the calculations
are terminated. Otherwise, the index \(k\) is incremented by \(p\), and new global optimization iteration is executed.
\par
The convergence conditions for the developed parallel algorithm are considered
in \cite{stronginGergelBarkalovParGO}, \cite{strSergGO}. Thus, in particular, when the convergence conditions are
satisfied, the parallel computations are non-redundant as compared to the serial method
when using up to \(2^N\) processors/cores (\(N\) is the dimensionality of the global optimization problem).

@@

problems (\ref{eq:oneDimTask}):

@@

        x_i':y'=y^i(x'),1\le i\le s,

@@

The possibility of asynchronous data transfer (the computational nodes process the obtained
data after receiving only) is an important feature of the proposed parallel computation scheme.

@@

\par
The architecture of the Globalizer system is presented in Figure \ref{fig:globalizerScheme}.
The system components are:
\begin{itemize}
  \item Block 0 consists of the procedures for computing the function values
  (criteria and constraints) of the optimization problem being solved; it is an external
  block with a predefined interface with respect to the Globalizer system and is provided by the user.
  \item Blocks 1-4 form the optimization subsystem and solve the global optimization
  problems (Block 1), nonlinear programming ones (Block 2), multicriterial optimization problems
  (Block 3), and general decision making problems (Block 4). The interaction between these
  components is organized in a hierarchical order --- the decision making problems are
  solved using the multicriterial optimization block, which, in turn, uses the nonlinear programming block, and so on.
  \item Block 5 is a subsystem for accumulating and processing the search information; it is one of the main subsystems – the size of search information may appear to be quite large and the efficiency of the global optimization methods depends on how completely the available search data are utilized.
  \item Block 6 contains the dimensionality reduction procedures based on the Peano space-filling curves; the reduced (one-dimensional) optimization problems are solved by the optimization blocks. This block also provides the interaction between the optimization blocks and the multidimensional optimization problem being solved.
  \item Block 7 provides tuning the parallel computation schemes in the Globalizer system subject to the employed computational system architecture (the number of the computational cores, the availability of shared and/or distributed memory, the availability of computational accelerators, etc.) and the applied global optimization methods.
  \item Block 8 is responsible for managing the parallel processes when performing the global optimization (determining the optimal configuration of parallel processes, distributing the processes between the computational units, balancing the computational loads, etc.).
  \item Block 9 is a managing subsystem, which controls the parallel computations when solving the global optimization problems.
  \item Block 10 provides the dialog interaction with the users for stating the optimization problem, adjusting the system parameters (if necessary), and visualizing and presenting the global optimization results.
  \item Block 11 is a set of tools for visualizing and presenting the global optimization results; the availability of the tools for visual demonstration of the computational results enables the user to perform an efficient control over the global optimization process.
\end{itemize}
\par
The Globalizer system comes in two different versions:
\begin{itemize}
  \item Globalizer Pro is a research version for making various large-scale numerical experiments on the high-performance computational systems to investigate the new parallel global optimization methods being developed.
  \item Globalizer Lite is a limited version for solving the global optimization problems with a medium complexity level, and is accessible for free usage.
\end{itemize}

\section{Results of Numerical Experiments}
\label{sec:experiments}
The numerical experiments were carried out using the Lobachevsky supercomputer at
the State University of Nizhni Novgorod (the operation system – CentOS 6.4, the
supercomputer management system – SLURM). Each supercomputer node had 2 Intel Sandy Bridge
E5-2660 processors, 2.2 GHz, with 64 GB RAM. Each processor had 8 cores (i. e. total
16 CPU cores were available at each node). In order to generate the executable program code, the Intel C++ 14.0.2 compiler was used.
\par
A number of computational nodes had Intel Xeon Phi 5110P processors, each containing 60 cores
(total 240 threads). In addition, the supercomputer included the computational nodes equipped
with two NVIDIA Kepler K20X GPUs, each providing 14 stream multiprocessors (total 2688 CUDA-cores). In order to generate the executable code, the CUDA Toolkit 6.0 was used.
\par
In the performed experiments the optimization problems generated by the GKLS-generator
\cite{gavianoKvasovLeraSergeev2003} were used. This generator allows obtaining the multiextremal optimization
problems with \textit{a priori} known properties: the number of local minima, the sizes of the attractors of these ones, the global minimum point, the function value at this point, etc. In order to simulate the computational costs that are inherent in the applied optimization problems, in all experiments computing the objective function was intentionally complicated by the auxiliary computations not altering the values of the function and the position of its minima.
\par
In all the executed experiments, 100 global optimization problems generated randomly by the GKLS-generator were solved. The results of experiments were averaged over the number of the solved optimization problems.

\subsection{Comparison with over methods}
The results of numerical comparison of three serial algorithms --- DIRECT
\cite{Jones}, DIRECT\textit{l} \cite{Gablonsky}, and MAGS from Section \ref{subsec:corepar} --- are presented below.

The numerical comparison was carried out by using the function classes Simple and Hard
of the dimensions 4 and 5. The results of numerical experiments for DIRECT and
DIRECT\textit{l} are taken from \cite{sergeyevKvasov2006}. According to \cite{sergeyevKvasov2006}, the global minimizer \(y^*\)
was considered to be found if the algorithm generated a trial point \(y^k\) inside
a hyperinterval with a vertex \(y^*\) and
\begin{displaymath}
    |y^k(i)-y^*(i)|\le \sqrt[N]{\Delta}(b(i)-a(i)), i\le i\le N,
\end{displaymath}
where \(N\) is the problem dimensionality, $a$ and $b$ are the borders of the search domain \(D\)
\footnote{In the case of MAGS, the stop condition was different slightly --- the
global minimizer \(y^*\) was considered to be found, if MAGS generates a trial
point \(y^k\)  in the \(\delta\)-vicinity of the global minimum, i. e. \(||y^k-y^*||\le\delta\).
The size of the vicinity was selected  as \(\delta=\left\|b-a\right\|\sqrt[N]{\Delta}\).}.
